---
title: "Final Report: Your Title Here"
author: "Don Francisco"
date: "November 22, 2017"
output:
    github_document: default
    word_document: default
bibliography: references.bib
csl: bioinformatics.csl
---

# Introduction

Add about 2-3 pages here. Across the whole manuscript, you should cite at least 20 peer reviewed articles.

Breast cancer affects affects one in eight women in their lifetime and according to the American Cancer Society, it is the second leading cause of cancer death in women with only lung cancer killing more women each year. Over the past twenty years $5.5 billion has been dedicated to breast cancer research but the origins of a majority of breast cancer cases remain unknown [@madigan1995proportion]. Part of what is known is that environmental factors as well as genetics play a role in disease development but which environmental factors and the extent of their influence is still being explored [@urbaniak2016microbiota]; diet, age and genetic disposition are established risk factors [@xuan2014microbial]. In fact, there are many epidemiological evidences and research studies that have suggested that diet plays an important role in breast cancer prevention and progression [@aragon2014modification]. Specifically, there are recent studies that suggest that eating fermented foods such as probiotic yogurt play a role in providing protection [@lakritz2014beneficial]. The recent appreciation of the influential role that microbiota has on human health and disease has led to the investigation of if microbes play a role in breast cancers and if they do, to what extent [@xuan2014microbial].

# Methods

## Study design

Add about half a page here. In this section instead of first person (I/we), use the authors of the paper you selected, since you'll just be describing what they did, based on the methods in their paper.

## Sample origin and sequencing

Add about half a page here. In this section instead of first person (I/we), use the authors of the paper you selected, since you'll just be describing what they did, based on the methods in their paper.

## Computational

These are the methods you used to do your bioinformatic analyses analyses. Should probably be between 0.5 and 1 pages. At a very minimum should include citations for DADA2 and phyloseq if you are doing an amplicon study, or other citations as appropriate.

# Results

## Subsections are ok in the results section too

```{r your-code-here}
# Add code chunks as needed for your analyses
# For most analyses, I would recommend splitting the
# intensive computational part into a seperate R script file
# and then just load your libraries and the data object here
# using the `load()` function. Ask me for clarification if this is
# unclear.
```

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")

# These are the primary packages well use to clean and analyze the data
# this package needs to be installed from bioconductor -- it's not on CRAN
# see info here: https://benjjneb.github.io/dada2/dada-installation.html
library("dada2")

# This to export a fasta of our final denoised sequence variants
library("seqinr")

# To install this you have to install from GitHub
# See more info here: https://github.com/leffj/mctoolsr
# run this -- install.packages("devtools")
# and then this -- devtools::install_github("leffj/mctoolsr")
library("mctoolsr")

# And this to visualize our results
# it also needs to be installed from bioconductor
library("phyloseq")

# This loads my separate R script file that processed the data
load("output/phyloseq_obj.RData")
```
In addition to a minimum of 5-10 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion. 2-3 pages.

# Discussion

Add around 3-4 pages interpreting your results and considering future directions one might take in analyzing these data.

# Sources Cited
